---
title: "Visualizing spatial data"
author: Alec Robitaille
date: 'May 4 2020 (updated: `r format(Sys.Date(), "%B %d %Y")`)'
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: todo
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE,
											highlight = TRUE,
											eval = TRUE)

tutorial_options(exercise.eval = FALSE)
```


## Packages

```{r packages}
library(mapview)
library(data.table)
library(sf)
library(ggplot2)
```

## Data
```{r data, exercise = TRUE}
## Load breweries data from mapview
data(breweries)

## Convert to data.table for example
# Grab the coordinates from the sf object and cast the matrix as a data.table
xy <- data.table(st_coordinates(breweries))

# Cast the sf object as a data.table
dtbrew <- data.table(breweries)

# Soft code coordinate column names (for later)
coordcols <- c('X', 'Y')

# Add the xy to the data.table
dtbrew[, (coordcols) := xy]
```


## Plotting sf objects

```{r plot1, exercise = TRUE}
plot(breweries)
```


Why so many subplots?

```{r plot2, exercise = TRUE}
## Set it universally as a default
options(sf_max.plot = 1)

# Or individually for each plot
# eg. 1
plot(breweries, max.plot = 1)

# eg. 4
plot(breweries, max.plot = 4)


## Or specify directly the column name to plot
plot(breweries['name'])
```


## Plotting sf objects with ggplot
Use `geom_sf`!

```{r, eval = FALSE}
ggplot(breweries) + 
	geom_sf()
```


Explore the breweries data and modify the aesthetics. 

```{r plot3, exercise = TRUE}
ggplot(breweries, aes()) + 
	geom_sf()
```

```{r plot3-solution}
ggplot(breweries, aes(color = brewery)) + 
	geom_sf()
```



## Plotting data.table spatial objects

data.tables can be plotted using ggplot with regular geoms. 

Plot the data.table version of the breweries data.

```{r plot4, exercise = TRUE}
dtbrew

??ggplot
```

```{r plot4-solution}
ggplot(dtbrew, aes(X, Y)) +
	geom_point()
```



## TODO:

mapview with sf
mapview with data.table

show no basemap because of missing crs
st_crs
