---
title: "Visualizing spatial data"
author: Alec Robitaille
date: 'May 4 2020 (updated: `r format(Sys.Date(), "%B %d %Y")`)'
output:
  learnr::tutorial:
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
description: todo
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE,
											highlight = TRUE,
											eval = TRUE)

tutorial_options(exercise.eval = FALSE)


library(mapview)
library(data.table)
library(sf)
library(ggplot2)

## Load breweries data from mapview
data(breweries)

## Convert to data.table for example
# Grab the coordinates from the sf object and cast the matrix as a data.table
xy <- data.table(st_coordinates(breweries))

# Cast the sf object as a data.table
dtbrew <- data.table(breweries)

# Soft code coordinate column names (for later)
coordcols <- c('X', 'Y')

# Add the xy to the data.table
dtbrew[, (coordcols) := xy]
```


## Packages

```{r packages}
library(mapview)
library(data.table)
library(sf)
library(ggplot2)
```

## Data
```{r data, exercise = TRUE}
## Load breweries data from mapview
data(breweries)

## Convert to data.table for example
# Grab the coordinates from the sf object and cast the matrix as a data.table
xy <- data.table(st_coordinates(breweries))

# Cast the sf object as a data.table
dtbrew <- data.table(breweries)

# Soft code coordinate column names (for later)
coordcols <- c('X', 'Y')

# Add the xy to the data.table
dtbrew[, (coordcols) := xy]
```


## Plotting: sf 

```{r plot1, exercise = TRUE}
plot(breweries)
```


Why so many subplots?

```{r plot2, exercise = TRUE}
## Set it universally as a default
options(sf_max.plot = 1)

# Or individually for each plot
# eg. 1
plot(breweries, max.plot = 1)

# eg. 4
plot(breweries, max.plot = 4)


## Or specify directly the column name to plot
plot(breweries['brewery'])
```


## Plotting: sf + ggplot
Use `geom_sf`! 
 
```{r, eval = FALSE}
ggplot(breweries) + 
	geom_sf()
```


Explore the breweries data and modify the aesthetics. 

```{r plot3, exercise = TRUE}


```

```{r plot3-solution}
ggplot(breweries, aes(color = brewery)) + 
	geom_sf()
```



## Plotting: data.table + ggplot

data.tables can be plotted using ggplot with regular geoms. 

Plot the data.table version of the breweries data.

```{r plot4, exercise = TRUE}
# Recall the data.table object is: 
dtbrew
```

```{r plot4-solution}
ggplot(dtbrew, aes(X, Y)) +
	geom_point()
```



## Interactive plotting: sf + mapview
[`mapview`](https://r-spatial.github.io/mapview/index.html) uses `leaflet` to generate interactive plots in R. The emphasis is on quick, exploratory plots and not publication figures. 

It supports plotting objects of class:

* sf
* data.frame / data.table
* raster
* stars
* sp
* satellite


Use the `mapview` function to plot the sf object. 
```{r map1, exercise = TRUE}

```

```{r map1-solution}
mapview(breweries)
```

## Interactive plotting: data.table + mapview

### 
`mapview` can also plot data.frames and data.tables. 

```{r map2, exercise = TRUE}
mapview(dtbrew)
```

###
Ok... we'll need to set some arguments.

* `xcol`
* `ycol`

```{r map3, exercise = TRUE}

```

```{r map3-solution}
mapview(dtbrew, xcol = 'X', ycol = 'Y')
```

### 
Close! But where did our basemap go?

We need to set the `crs`. Since our data comes from an `sf` object, we can use it's CRS information. 

```{r map4, exercise = TRUE}
st_crs(breweries)
```

```{r map4-solution}
mapview(dtbrew,
				xcol = 'X',
				ycol = 'Y',
				crs = st_crs(breweries))
```


###
Lastly... what about colors? 

Color the points by the brewery name using the `zcol` argument. 

```{r map5, exercise = TRUE}

```

```{r map5-solution}
mapview(
	dtbrew,
	xcol = 'X',
	ycol = 'Y',
	crs = st_crs(breweries),
	zcol = 'brewery'
)
```


###
On your own time, check customizing [popups](https://r-spatial.github.io/mapview/articles/articles/mapview_04-popups.html)

```{r map6, exercise = TRUE}
mapview(
	dtbrew,
	xcol = 'X',
	ycol = 'Y',
	crs = st_crs(breweries),
	zcol = 'brewery'
)
```


##
Note: `mapview` is getting faster all the time but it is still going to be limited when you making interactive plots of large datasets or large rasters. 
mapview with sf
mapview with data.table

show no basemap because of missing crs
st_crs
