---
title: Troubleshooting
subtitle: and problem solving in R
author: Alec L. Robitaille
institute:  Wildlife Evolutionary Ecology Lab <br> Memorial University
date: "2022-04-18"
output: 
  xaringan::moon_reader:
    nature:
      beforeInit: "2022-04-18-troubleshooting.js"
      highlightStyle: agate
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
    css: ["2022-04-18-troubleshooting-xaringan-themer.css"]
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
base_color <- "#b68e7d"
lighter_base <- lighten_color(base_color, 0.35)
h4_color <- darken_color(lighter_base, 0.2)

regular <- "1rem"
small <- "0.75rem"
h1 <- "2.25rem"
h2 <- "2rem"
h3 <- "1.3rem"

code_color <- darken_color(base_color, 0.7)

style_mono_dark(
	outfile = '2022-04-18-troubleshooting-xaringan-themer.css',
  base_color = base_color,
  link_color = lighter_base,
  code_inline_color = lighter_base,
	left_column_selected_color = apply_alpha(base_color, 0.7),
  code_inline_background_color = code_color,

  text_font_size = "1.5rem",
  header_h3_font_size = "0.75rem",
  extra_css = list(
		".remark-slide-content h4" = list("font-size" = regular, "color" = h4_color),
		".left-column" = list("width" = "20%", 'font-size' = regular,
													"height" = '95%'
													# 'border-right' = '0.1px solid var(--link-color)'
													),
		".hljs-agate .hljs" = list("background" = code_color, "color" = lighter_base),
		".panelset" = list("font-size" = regular),
		".medium" = list("font-size" = regular),
		".small" = list("font-size" = small),
		".footnote" = list("font-size" = small,  "left" = '75%'),
		".title-slide h1" = list("font-size" = h1),
		".title-slide h2" = list("font-size" = h2),
		".title-slide h3" = list("font-size" = h3),
		"img" = list("box-shadow" = "2px 2px 5px var(--link-color)")
	),
  header_font_google = google_font("Roboto Slab", "200"),
  text_font_google = google_font("Roboto Slab", "300"),
  code_font_google   = google_font("Fira Mono")
)
xaringanExtra::use_panelset()
```



.left-column[
<br><br><br><br><br><br>
### **Discussion**
### Prevention
### Simplifying
### and other strategies
### Resources
]


---

.left-column[
<br><br><br><br><br><br>
### **Discussion**
### Prevention
### Simplifying
### and other strategies
### Resources
]


#### What are your strategies for


...troubleshooting errors?

...problem solving?

...learning a new package?

...starting a new analysis?



---

.left-column[
<br><br><br><br><br><br>
### Discussion
### **Prevention**
### Simplifying
### and other strategies
### Resources
]

#### First, some preventative measures

Adjust RStudio settings [[1]](https://csgillespie.github.io/efficientR/set-up.html#rstudio-options)

Use an RStudio project, relative file paths

Keep R, R packages, and OS updated


---

.left-column[
<br><br><br><br><br><br>
### Discussion
### **Prevention**
### Simplifying
### and other strategies
### Resources
]

#### Extras

Track package versions with `renv`

Reproducible workflows with `targets`

Use Git

???
renv tracks package versions for each project, so global updates to packages 
won't break eg. older projects

Git helps with the feedback cycle of coding -> tracked changes -> commit -> coding
Also, Git lets you create branches to explore alternatives, without 
breaking your current working versions

A workflow defined in targets ensures steps are only rerun when they are outdated 
and helps branching by eg. individuals or study sites


---


.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

####  Overview

Simplify the complexity of the task

...to identify source of the error

...to clarify steps of the analysis

...to learn new functions or packages


???

Tasks: troubleshooting, learning a new package, starting a new analysis...

We'll first look at making minimal data, then using that data to
learn a new package, then finally taking a complex script + data and
reducing it to a minimum example


---


.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

#### Data

Goal: minimal and representative data

Option 1 - available example data

Option 2 - build your own

???

Representative: retains variables, types and grouping levels of original data

---

.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

#### Available data

.panelset[
.panel[.panel-name[Penguins]

```{r}
library(palmerpenguins)
data(penguins)
nrow(penguins)
colnames(penguins)
```
]
.panel[.panel-name[Cars]
```{r}
data(mtcars)
nrow(mtcars)
colnames(mtcars)
```
]
.panel[.panel-name[Etc]
```{r, eval = FALSE}
# Run data() to see a list of available data
data()

# for spatial data, see the spData package
data(package = 'spData')
```
]
]
---

.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

#### Exercise: build your own data

1) Take your real data

2) List  
-columns  
-data types  
-units  
-limits  
-factors or levels

3) Generate a representative and minimal fake data

.footnote[
[`R/simplify-data.R`](https://github.com/robitalec/workshops/tree/master/2022-04-18-troubleshooting/R/simplify-data.R)
]



---
.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

#### Learning a new package

Use a simple dataset

to focus on learning functions 

without complexity of real data or specific analyses


---
.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

#### Exercise: learn a new package

1) Use a minimal dataset

2) Open package website or documentation

3) Fill the [template](https://github.com/robitalec/workshops/tree/master/2022-04-18-troubleshooting/R/learn-package-template.R) (see the [example](https://github.com/robitalec/workshops/tree/master/2022-04-18-troubleshooting/R/learn-package-chk.R))

4) Learn a few functions!

.footnote[
Pick any package or one from this list:  
-[`ggplot2`](https://ggplot2.tidyverse.org/)  
-[`sf`](https://r-spatial.github.io/sf/)  
-[`chk`](https://poissonconsulting.github.io/chk/)  
-[`anytime`](https://eddelbuettel.github.io/anytime/)    
-[`patchwork`](https://patchwork.data-imaginist.com/)  
]



---
.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

#### Troubleshooting





---
.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### **Simplifying**
### and other strategies
### Resources
]

#### Exercise: troubleshooting

Given a complex script (e.g.  [1](https://github.com/wildlifeevoeco/SocCaribou/blob/master/scripts/2-HomeRangeAnalysis.R), [2](https://github.com/robitalec/ScaleInMultilayerNetworks/blob/master/scripts/03-temporal-layers.R), [3](https://github.com/wildlifeevoeco/study-area-figs/blob/master/R/10-riding-mountain-prep.R), [4](https://github.com/robitalec/spatsoc-application-paper/blob/master/scripts/figure-2-group-pts.R)),  
build a reprex


1) Start a new script

2) Only load essential packages

3) Use a minimal dataset

4) Isolate the erroring function

5) Reproduce the error  
or find your mistake


.footnote[
[`reprex` vignette](https://reprex.tidyverse.org/articles/reprex-dos-and-donts.html)  
[SO reprex how to](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example/16532098)
]

???
Ideally, bring your own script

Say you hit an error halfway through some complex analysis... 

`reprex_selection`

---

.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### Simplifying
### **and other strategies**
### Resources
]

Explore the traceback [[1]](https://adv-r.hadley.nz/debugging.html#traceback)

Read (and reread) documentation

Restart R, clear the environment

Update R, R packages, OS

Check your .Rprofile, .Renviron
[[2]](https://csgillespie.github.io/efficientR/set-up.html#an-overview-of-rs-startup-files=), [[3]](https://usethis.r-lib.org/reference/edit.html)

GitHub Issues [(e.g. [4])](https://github.com/rstudio/bookdown/issues/1263)  
Stack Overflow [(e.g. [5])](https://stackoverflow.com/questions/42149427/group-membership-by-pairwise-criteria-matching)

Ask for help! 



---

.left-column[
<br><br><br><br><br><br>
### Discussion
### Prevention
### Simplifying
### and other strategies
### **Resources**
]

-[`reprex`: introduction](https://reprex.tidyverse.org/articles/reprex-dos-and-donts.html)  
-[SO: Making a great R reproducible example](https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example/16532098)  
-[`reprex`: datapasta](https://reprex.tidyverse.org/articles/datapasta-reprex.html)


-[Advanced R: Debugging](https://adv-r.hadley.nz/debugging.html0)


-[Description of startup files](https://csgillespie.github.io/efficientR/set-up.html#an-overview-of-rs-startup-files=)  
-[Editing startup files](https://usethis.r-lib.org/reference/edit.html)

-[`targets` manual](https://books.ropensci.org/targets/)  
-[`renv` documentation](https://rstudio.github.io/renv/)

-[Efficient R: Version control](https://csgillespie.github.io/efficientR/collaboration.html#version-control)  
-[Happy Git with R](https://happygitwithr.com/)
