---
title: "Introduction to collar data"
author: Alec Robitaille
date: 'May 4 2020 (updated: `r format(Sys.Date(), "%B %d %Y")`)'
output:
  learnr::tutorial:
    allow_skip: false
runtime: shiny_prerendered
description: >
  This tutorial introduces working with collar data in R. Since collar data is often obtained in a spreadsheet format, we will assume a CSV input file format. 
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE,
											highlight = TRUE)

tutorial_options(exercise.eval = FALSE)
```


## Packages

```{r packages}
library(mapview)
library(data.table)
```



# TODO:
* read in csv
* read in many csvs and combine
* ordering data
* working by individual
* converting between data.table and sf 
* plotting trajectories with ggplot (time ordered)
* gold lines


introduce these concepts using the basic math / functions
there are helper packages (list them) that assist in this

difference in time between relocations
difference in x, y between relocations
step length

caution when working with collar data
* time zones (use UTC where possible)
* missed fixes
* irregular fix rates
* there is error to GPS fixes!  topography
* biological relevance
