---
title: "Reading and writing spatial data in R"
output:
  learnr::tutorial:
    allow_skip: false
runtime: shiny_prerendered
description: >
  This tutorial demonstrates reading and writing spatial data in R, and has a basic comparison of different spatial file formats. 
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE, 
                      highlight = TRUE)

tutorial_options(exercise.eval = FALSE)
```


<!--

  allow_retry = TRUE
  random_answer_order

echo = FALSE

quiz (
questions..
)
post_message

-->

## Packages

```{r packages}
library(sf)
library(mapview)
library(data.table)
```

## Example data

```{r data}
# Load breweries data from mapview
data(breweries)

# Have a look!
plot(breweries, max.plot = 1)
```

```{r map}
# Or interactive:
mapview(breweries)

```


## Spatial file formats

```{r, echo = FALSE}
knitr::kable(fread('../../extdata/data-formats.csv', header = TRUE, col.names = c('File format', '')))
```

Note: We are ignoring raster-based file formats for now. 


### Are these available in R?

```{r drivers, exercise = TRUE, exercise.eval = FALSE}
# In the sf package, the function st_drivers() gives us all available file formats
drivers <- st_drivers()

# Explore the drivers to find a couple of the above file formats
# If we convert to a data.table we can filter easily
data.table(drivers)[name == 'GPKG']
```


## Writing spatial data

Since we have the `breweries` data loaded, let's write it out using `sf::st_write`. 

```{r write, echo = TRUE, eval = FALSE}
# Write to a GeoPackage
st_write(breweries, 'output/breweries.gpkg')
```

Some arguments from the `st_write` function:

* 'obj': the sf or sfc object
* 'dsn': data source name.
* 'layer': layer name. if layer is missing, the basename is used (eg. "breweries" above).
* 'driver': guesses from the dsn, or passed explicitly here.
* 'overwrite': default FALSE. set it to TRUE when are going to be updating the file (be careful!). 

The arguments 'dsn' and 'layer' depend on which drive/file format is used. This can be a bit confusing at times - usually we can just provide the full path with the filename and it should work! 

### Quiz!

Write out the breweries to a GeoPKG, Shapefile, and GeoJSON. Check your solution by hitting "Run" or look at the solution. 


```{r writequiz, exercise = TRUE, exercise.eval = TRUE}
# Create an output folder
dir.create('output')

# Write to a GeoPackage


# Write to a Shapefile


# Write to a GeoJSON


# Check for yourself!
dir('output/', full.names = TRUE)
```


```{r writequiz-solution}
# Create an output folder
dir.create('output')

# Write to a GeoPackage
st_write(breweries, 'output/breweries.gpkg')

# Write to a Shapefile
st_write(breweries, 'output/breweries.shp')

# Write to a GeoJSON
st_write(breweries, 'output/breweries.GeoJSON')

# Check for yourself!
dir('output/', full.names = TRUE)
```


#### Note: A warning was generated when writing to shapefile

```
Warning in abbreviate_shapefile_names(obj): Field names abbreviated
for ESRI Shapefile driver
```

One disadvantage to using shapefiles is they limit attribute names to 10 characters. 

Let's see what happened:

```{r limnames, eval = TRUE, echo = TRUE}
colnames(breweries)
colnames(st_read('output/breweries.shp', quiet = TRUE))
```


Thankfully we were warned... Otherwise we might be confused by "nmbr_f_" and "nmbr_s_" in our output. 
