---
title: 'data.table or: How I Learned To Stop For-Looping and Love With I Do J By Group'
author: Alec Robitaille 
date: "2017-11-16"
output: 
  xaringan::moon_reader:
    nature:
      highlightStyle: github
      beforeInit: 'macros.js'
      highlightLines: true
      slideNumberFormat: |
        <div class='progress-bar-container'>
          <div class='progress-bar' style='width: calc(%current% / %total% * 100%);'>
          </div>
        </div>`
    css: [metropolis, metropolis-fonts, 'extra.css']
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE,
                      fig.align='center',
                      message = FALSE, warning = FALSE)

```

```{r, echo = FALSE, message = FALSE}
library(data.table)
library(ggplot2)
diamonds <- fread('../input/diamonds.csv')
```

# What is wrong with my for loops?

for loops (*in R*) are inefficient, especially when growing objects iteratively. 

they can be improved by pre-allocating the vectors or lists, if we know the expected size of the final objects. 

regardless, there are more concise and efficient ways of writing R (both in number of keystrokes and processing efficiency)



```{r, echo = FALSE}
DF <- data.frame(n = runif(1e5, 1, 5),
                 grp = seq(1:100))
# for(i in seq_along(unique(DF$grp))){
#   DF[DF$grp == i,] + DF[DF$grp == i+1,]
# }
data(diamonds)
```

---
class:middle, center, clear, review

`DT[i, j, by]`

which reads: 

> with I do J by group

???
whenever you want to write a data.table expression, repeat these words...

---

# Modify by reference

One of data.table's main driving philosophies is to modify objects *by reference*. 

This avoids the memory cost of assigning and reassigning objects by changing them in place. 

This also means that methods on small tables translate better to large ones. 

---

# Creating a data.table



.pull-left[
data.tables can go anywhere data.frames go because they share the data.frame class:
```{r, echo = TRUE}
diamonds <- data.frame(diamonds)

setDT(diamonds)

class(diamonds)
```
]

.pull-right[
Function  | Second Header
:------------ | :------------
`setDT`  | coerce by reference lists or DFs
`as.data.table`  | cast to data.table (requires assignment)
`data.table`  | general constructor
]



???
differentiate setDT and data.table and as.data.table
note that this isn't the only way
if it is used though, it only needs to be run
once. stick it on a line and move on

---

# data.table "i"

forget `base::subset` and its weird issues


```{r, eval = FALSE, echo = TRUE}
diamonds[cut == 'Very Good']
```
```{r}
diamonds[cut == 'Very Good'][1:5]
```
```{r, eval = FALSE, echo = TRUE}
diamonds[carat < 0.24 & color != 'J']
```

```{r}
diamonds[carat < 0.24 & color != 'J'][1:5]
```

---
# data.table "i": ordering the rows

```{r, eval = FALSE, echo = TRUE}
diamonds[order(price)]
```

```{r}
diamonds[order(price)][1:5]
```

```{r, eval = FALSE, echo = TRUE}
diamonds[order(depth, price, carat)]
```

```{r}
diamonds[order(depth, price, carat)][1:5]
```

???
a quick note about non-standard evaluation
fill more actual description about what non-standard eval is
notice how in the previous example, we used the column names directly?

so easy, so clean..

---
# data.table  "j"

this is where we can select, compute on, add or delete columns. 

the only requirement for j is to return a list, list of lists or data.frame/data.table. ie: no SpatialPolygons, etc. [1]


```{r, echo = TRUE}
diamonds[, sum(price)]
```
```{r, echo = TRUE}
diamonds[, .(sum(price))]
```

.footnote[
[1] You can store complex data in columns if they are within a list. See [here](http://brooksandrew.github.io/simpleblog/articles/advanced-data-table/#columns-of-lists) for an example.
]

---
# data.table  "j"

```{r, echo = TRUE}
diamonds[, .(sumPrice = sum(price))]
```

```{r, echo = TRUE}
diamonds[, .(sumPrice = sum(price),
             meanCarat = mean(carat))]
```

???
j is the real workhorse.
without intermediate output changed my life
highlight the differences between these three lines

---
# :=  add/modify columns 

For one column:

```{r, echo=TRUE}
diamonds[, meanPrice := mean(price)]
```

```{r}
diamonds[1:3, .(carat, cut, color, clarity, depth, meanPrice)]
```


---
# :=  add/modify columns 

For multiple columns:

```{r, echo=TRUE}
diamonds[, c('meanPrice', 'meanCarat')  := .(mean(price), mean(carat))]
```

Or with a list of column names, wrap with brackets. 
```{r}
outcols <- c('meanPrice', 'meanCarat')
diamonds[, (outcols)  := .(mean(price), mean(carat))]
```

---

# := remove columns
Our example data:
```{r}
diamonds[1:3, .(carat, cut, color, clarity, depth, meanPrice)]
```

Remove the meanPrice column
```{r, echo = TRUE}
diamonds[, meanPrice := NULL]
```

```{r}
diamonds[1:3, .(carat, cut, color, clarity, depth)]
```

---

# := remove columns

Remove multiple columns
```{r, echo = TRUE, eval = FALSE}
diamonds[, c('carat', 'cut') := NULL]
```

```{r}
diamonds[1:3, .(color, clarity, depth)]
```


---

# data.table "by"

## Single column
```{r, echo = TRUE}
diamonds[, mean(price), by = color]
```

## Multiple columns
```{r, echo = TRUE, eval = FALSE}
diamonds[, mean(price), by = .(color, clarity)]
diamonds[, mean(price), by = c('color', 'clarity')]
diamonds[, mean(price), by = color:clarity]
```

???
how do we fix this column name?
instant pop quiz review
alternatively, how can we add it as a column


highlight the clarity and unambiguous nature of this syntax
we aren't using any strange numbers that have no meaning


---
#  data.table "by"

## Group by a conditional statement

```{r, echo = TRUE}
diamonds[, mean(price), by = price > 500]
```

## Group by a list of columns 

```{r, eval = FALSE, echo = TRUE}
cols <- c('color', 'clarity')

diamonds[, mean(price), by = cols]

```

---

# Chaining expressions 

```{r, echo = TRUE, eval = FALSE}
diamonds[, mnPrice := mean(price), by = cut][, dif := price - mnPrice]
```
```{r}
diamonds[, mnPrice := mean(price), by = cut][, dif := price - mnPrice][sample(.N, 5), .(price, mnPrice, dif)]
```

???
order locs by datetime then calculate step length

---

# data.table special symbols

`.SD`
- **S**ubset of the **D**ata.table

`.BY`
- list of items in the by

`.I`
- row numbers in the data.table

`.GRP`
- group counter

`.N`
- number of rows (in each group)


---
# data.table .SD

.left-tight[
.SD is the list of columns in the data.table and represents the subset of the data.table when grouping. 

.SDcols informs data.table that only a subset of the columns are needed for a function. 

(note that .SDcols doesn't accept the .(col1, col2) syntax)

]

.right-wide[

```{r, eval = TRUE, echo = TRUE}
diamonds[, lapply(.SD, mean), .SDcols = c('price', 'carat')]
```

```{r, eval = TRUE, echo = TRUE}
diamonds[, lapply(.SD, mean), 
         .SDcols = c('price', 'carat'), 
         by = color]
```
]

---

# data.table .N

.pull-left[
## How many rows?
```{r, eval = TRUE, echo = TRUE}
diamonds[, .N]
```

]

.pull-right[
## In a group?
```{r, eval = FALSE, echo = TRUE}
diamonds[, .N, by = .(color, cut)]
```

```{r, eval = TRUE}
diamonds[, .N, by = .(color, cut)][1:5]
```

]

???
recently juliana and I were running into an issue with
multiple locs for the same individual at the same time
this caused a major issue for

---

# data.table .BY


## For plotting:

```{r, eval = FALSE, echo = TRUE}
diamonds[, print(qplot(price, main = as.character(.BY))), by = color]
```

## Conditional statements on the value of the group

```{r, eval = FALSE, echo = TRUE}
bestCuts <- c('Ideal', 'Premium', 'Very Good')

diamonds[, .(adjustedMeanPrice = if(.BY %in% bestCuts){
  mean(price) + 200
} else {
  mean(price) - 200
}), by = cut]
```

## If using multiple by columns, simply select the .BY of interest

```{r, eval = FALSE, echo = TRUE}
diamonds[1:3, paste('the carat is...', .BY[1]),  by = .(carat, color)]
```

```{r, eval = FALSE, echo = TRUE}
diamonds[1:3, paste0(.BY[1], .BY[2]),  by = .(carat, color)]
```

# data.table .I

```{r, echo = TRUE}
diamonds[, rowN := .I]
```

```{r, echo = FALSE}
diamonds[1:5, .(cut, color, depth, price, rowN)]
```

```{r, echo = TRUE}
diamonds[, .I[which.max(price)], by = color]
```

# data.table .GRP
Group counter

```{r}
diamonds[, .GRP, by = color]
```

???
add grp as grp then use on grp later

---
class: important
# Good practices with data.table
build your expressions stepwise, ensuring each works before adding any columns to the DT. 


1. check that your subsetting is working
1. with subsetting check the j
1. with sub and j, check group 
1. then add the column, once you are sure the result is good. 


---
# Advanced methods

## Functions in data.table's i or by

```{r, echo = TRUE}
diamonds[price == max(price), .(carat, color, price, clarity, depth)]
```

```{r, echo = TRUE}
diamonds[sample(.N, 5), .(carat, color, price, clarity, depth)]
```

---

# Plenty of other functionality
* the amazing fread, fwrite
* setnames
* setcolorder
* shift
* merge
* dcast
* melt
* IDate, ITime

my favorite combo: lapply -> rbindlist



???
why you shouldn't use subset:
https://stackoverflow.com/questions/9860090/why-is-better-than-subset


---
class: clear

```{r, echo = TRUE, eval =FALSE}
### data.table Step Length ----
# Set columns
time.col <- 'datetime'
coord.cols <- c('EASTING', 'NORTHING')

# Create lag and dif column names
lag.cols <- paste('lag', coord.cols, sep = '')
difference.cols <- c('difX', 'difY')

lag.time.col <- paste0('lag', time.col)
dif.time.col <- paste0('dif', time.col)

# Use shift  to create lagged cols
locs[order(get(time.col)), (lag.cols) := shift(.SD, 1, NA, 'lag'),
     by = .(ANIMAL_ID, year),
     .SDcols = coord.cols]

# Find the difference squared between all points in each x,y separately
locs[, (difference.cols) := .((get(coord.cols[1]) - get(lag.cols[1])) ^2,
                              (get(coord.cols[2]) - get(lag.cols[2])) ^2)]

# Square root the summed difference for a simple step length
locs[, simpleStep := sqrt(rowSums(.SD)),
     .SDcols = difference.cols]
```

---

```{r, echo = TRUE, eval = FALSE}
## Delta Time
locs[order(get(time.col)), (lag.time.col) := shift(.SD, 1, NA, 'lag'), 
     by = .(ANIMAL_ID, year),
     .SDcols = time.col]

# difference in time in hours
locs[, (dif.time.col) := as.numeric(get(time.col) - get(lag.time.col), units = 'hours')]

# Simple step length divided by time difference
locs[, moveRate := simpleStep / (get(dif.time.col))]

```

---

class: clear
```{r, eval = FALSE, echo = TRUE}
# Drop columns leaving only needed
cols <- c('EASTING', 'NORTHING', 'blockByIDYear', 'ANIMAL_ID', 'block', 'HERD')

# and set all locs as observed
observed.locs <- locs[, ..cols][, observed := 1]

# Create identical for random with observed == 0
random.locs <- locs[, ..cols][, observed := 0]

# Generate an equivalent number of random points in vertices as observed (with encamped state)
# (updating the observed locs EASTING, NORTHING columns)
random.locs[, c('EASTING', 'NORTHING') :=
 as.data.table(spsample(vertices.95[vertices.95@data$id == .BY[[1]],],
      .N, iter = 100, type = 'random')@coords),
            by = blockByIDYear]
```
